---
title: "IODS week6 exercises"
author: "Sina Hulkkonen"
date: "12/8/2018"
output: html_document
---
*Describe the work you have done this week and summarize your learning.*

I've returned to my clinical work, so I didn't make it to the lectures and had minimal time for these exercises. I did all the DataCamp exercises.

At first, I copied the code from datawrangling exercise to get the long forms of these datasets:
```{r}
library(tidyr)
library(dplyr)
library(ggplot2)

#loading he datasets BPRS and RATS
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep  =" ", header = T)
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')
#both datasets include study IDs, and the participants have been divided into categories by treatment (BPRS) and group (RATS)
#converting categorical variables into factors
BPRS$treatment <- factor(BPRS$treatment)
RATS$Group <- factor(RATS$Group)
RATS$ID <- factor(RATS$ID)

##BPRS
#convert to long form
BPRSL <-  BPRS %>% gather(key = weeks, value = bprs, -treatment, -subject)
#extract the week number
BPRSL <-  BPRSL %>% dplyr::mutate(week = as.integer(substr(weeks,5,5)))

##RATS
# Convert data to long form & add Time variable
RATSL <- RATS %>%
  gather(key = WD, value = Weight, -ID, -Group) %>%
  mutate(Time = as.integer(substr(WD,3,4))) 
```

##Analysis for RATS (MABS chapter 8)
#Looking at the data I wrangled before
```{r}
summary(RATSL)
str(RATSL)
glimpse(RATSL)
```

The RATS data contains information on rats gaining weight on specific diet. The three groups were put on different diets, and each animal’s body weight (grams) was recorded repeatedly (approximately weekly, except in week seven when two recordings were taken) over a 9-week period.

#Graphical Displays of Longitudinal Data RATSL
```{r}
library(ggplot2)

# Draw the plot
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))

# Standardise the variable Weight
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(Weight_std = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()

# Draw the plot with scaled weights
ggplot(RATSL, aes(x = Time, y = Weight_std, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight_std), max(RATSL$Weight_std)))

```

Here in the first pictures we see, how the weights of the rats in each group developed by time. In the second picture, the weights of the rats are standardized, since they are different in the beginning. This enables us to compare the change they had over time.

#Summary Measure Analysis of Longitudinal Data for RATSL
```{r}
# Number of weeks, baseline (week 0) included
n <- RATSL$Time %>% unique() %>% length()

# Summary data with mean and standard error of bprs by treatment and week 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)

# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.5)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")

```

In this picture, we see summary of each group of the rats' weights mean (whiskers represent standard deviation) in different time points. It looks like all the rats gained weight compared to the starting point, the most in group 2.

#Applying the Summary Measure Approach
```{r}
# Create a summary data by treatment and subject with mean as the summary variable (ignoring baseline week 0).
RATSLS <- RATSL %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

glimpse(RATSLS)

# Draw a boxplot of the mean versus treatment
ggplot(RATSLS, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Time > 0")

#Cannot do T-test, since group has three levels, lets do chi-squred instead
chisq.test(RATSLS$Group, RATSLS$mean)

```

Here we see the mean weight in each group after the first time point. It seems that the rats weighed more in groups 2 and 3. The chi-squared p-value=0.3675 tells us that there is not significant change in the mean weght of the rats in different groups.

#Incorporating Pre-Treatment Outcome Values into the Summary Measure Approach
```{r}
# Add the baseline from the original data as a new variable to the summary data
RATSLS <- RATSLS %>%
  mutate(baseline = RATS$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = RATSLS)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)
```

Baseline measurements of the outcome variable in a longitudinal study are often correlated with the chosen summary measure and using such measures in the analysis can often lead to substantial gains in precision when used appropriately as a covariate in an analysis of covariance. Here we see, that te baseline (the weight of the rats in the beginning of the study) is correlated with mean weight gain in the follow-up. This means, that if the groups differ by weight in the beginning, the results might be biased.


##Analysis for BPRS (MABS chapter 9)
#Looking at the data I wrangled before
```{r}
summary(BPRSL)
str(BPRSL)
glimpse(BPRSL)
```
BPRS data contains information of 40 male subjects were randomly assigned to one of two treatment groups and each subject was rated on the brief psychiatric rating scale (BPRS) measured before treatment began (week 0) and then at weekly intervals for eight weeks. The BPRS assesses the level of 18 symptom constructs such as hostility, suspiciousness, hallucinations and grandiosity; each of these is rated from one (not present) to seven (extremely severe). The scale is used to evaluate patients suspected of having schizophrenia.

#Linear mixed effect models for repeated measures data
Please see the data wrangling part! I converted BPRS into longitudinal form previously.

#Fitting the Independence Model to the BPRSL Data
```{r}

#For this plot to succeed, I have t make new factors
BPRSL$treatment<-factor(BPRSL$treatment)
BPRSL$subject<-factor(BPRSL$subject)

# Plot the BPRSL data
#p<-ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) 
#p1<- p+geom_line(aes(linetype = treatment)) 
#p2<-p1+ scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1))
#p3<-p2+scale_y_continuous(name = "bprs (points)")
#p4<-p3+theme(legend.position = "top")
#help("geom_line")
#p4

p<-ggplot(BPRSL, aes(x = week, y = bprs, group = subject))
p1<- p+geom_line(aes(linetype = subject)) 
p2<-p1+ scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1))
p3<-p2+scale_y_continuous(name = "bprs (points)")
p4<-p3+theme(legend.position = "top")
p4

#ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +geom_line(aes(linetype = treatment)) + scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1))+scale_y_continuous(name = "bprs (points)")+theme(legend.position = "top")
```

Here we ignore that there are multiple observations belonging to one individual (independence model). For some reason, the command ends up in error. I was able t draw a plot where single individuals bprs scores are shown week by week.


#Fitting Linear Mixed Models to the BPRS Data
```{r}
# create a regression model RATS_reg
BPRS_reg <- lm(bprs~week + treatment, data=BPRSL)

# print out a summary of the model
summary(BPRS_reg)

summary(BPRSL)

help("pairs")
pairs(bprs~week, col=BPRSL$treatment, data=BPRSL)

```

Here we see that bprs points reduce by 2.27 when week increses by one (in one week, in other words) with significant p-value. On the other hand, bprs points seem to increase by 0.57 when treatment 2 is chosen over treatment 1, but this finding is not statistically significant (p=0.661). The scatterplot shows how the bprs points were divided for each treatment group in different time points.

I have to quit here. I ran into some difficulties here, and couldn't find answers quick enough. I think this'll do for me :)
```{r}
```

```{r}
```
