# IODS week5 RStudio exercise

Sina Hulkkonen
1/12/2018

*Describe the work you have done this week and summarize your learning.*

I did all the DataCamp exercises for this week. For the first time I didn't make it to the lecture, so I tried to manage myself with the exercises alone.

#1. Describing the data
```{r}
library(GGally)
library(dplyr)
library(corrplot)
library(tidyr)
human<-read.table("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/human2.txt", sep=",", header = TRUE)

#summary(human)

ggpairs(human)

# compute the correlation matrix and visualize it with corrplot
cor(human)%>%corrplot()
```
The 'human' dataset originates from the United Nations Development Programme. It contains information on education, birth and maternal mortality rates, life expectancy etc in different countries. In ggpairs plot, we see the distribution in each variable and how the variables correlate with each other. In correlation plot the size of the plot tells about how strong the correlation is, red dots representing negative and blue dots positive correlation. For example, maternal mortality and life expectancy have a strong negative correlation.

#2. Principal component analysis (PCA) on the not standardized human data
```{r}
# perform principal component analysis (with the SVD method)
pca_human <- prcomp(human)

#showing the variance captured by the components
summary(pca_human)

# draw a biplot of the principal component representation and the original variables
biplot(pca_human, choices = 1:2, cex = c(0.8, 1), col = c("grey40", "deeppink2"))
```
It looks like the component PC1 captures all the variance. This might be due the dataset is not standardized, and the variable variances may vary tremendously, which leads the PCA to think that the variable with the most variance is the most important (although it might not be). This is why the plot is ugly, I think.

#3. Principal component analysis (PCA) on the standardized human data
```{r}
human_std <- scale(human)

# perform principal component analysis (with the SVD method)
pca_human_std <- prcomp(human_std)

#showing the variance captured by the components
summary(pca_human_std)

# draw a biplot of the principal component representation and the original variables
biplot(pca_human_std, choices = 1:2, cex = c(0.8, 1), col = c("grey40", "deeppink2"))
```
Please see the interpretation of the unstandardized data PCA in the previous paraghaph. Now the variance of principal components divides more beatofully, with PC1 capturing 53.6% of the variance and the others less. The plot is prettier. It shows that the arrows form three groups: maternal mortality and adolescent birth rate, percentage of women represantatives in parliament and women:men labour ratio, and the rest of the variables (life expectancy, education rate and ratio and Gross National Income per capita). This tells that they actually might be related and measure a similar feature of the population or country.

#4. Interpretation of the previous plot
It is logical that where there are more women involved in working life, they are also more women as parliament representatives. Also, we do know that women give birth younger in developing countries with poorer healthcare systems. In countries where adolescents give birth, there seems to be higher maternal mortality.

#5. Tea time!
```{r}
#install.packages("FactoMineR")
library(FactoMineR)
library(ggplot2)

#getting tea dataset
data("tea")

# look at the summaries and structure of the data
summary(tea)
str(tea)

#too many variables to handle, so I decided to make dataset tea_time
keep_columns <- c("Tea", "How", "how", "sugar", "where", "lunch")

# select the 'keep_columns' to create a new dataset
tea_time <- select(tea, one_of(keep_columns))

# look at the summaries and structure of the data
summary(tea_time)
str(tea_time)

# visualize the dataset
gather(tea_time) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar()+theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))

# multiple correspondence analysis
mca <- MCA(tea_time, graph = FALSE)

# summary of the model
summary(mca)

# visualize MCA
plot(mca, invisible=c("ind"), habillage="quali")

# visualize MCA
plot(mca, invisible=c("var"), habillage="quali")

```
I decided to keep only some variables in the dataset, since there were too many in the original dataset. MCA is run for my dataset. I printed out both variables and individuals. The first plot (individuals) is informative and tells us which variable factors represent each other. For example, individuals that buy unpacked tea are likely to buy it from tea shop (see the factors levels nearby each other in the plot).